Protecting causal effects

### 2016-06-08 identify/identify.tex
adds initial writup of contextual-bandit and identifiability problems

\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{icml2016stylefiles/icml2016} 
\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\setlength{\marginparwidth}{10ex}%added by Yifan so that the todos can fit in the margin
%\usepackage[obeyFinal]{todonotes}
\usepackage[disable]{todonotes}

@@

\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}

@@

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Causal Bandits}

@@

\twocolumn[
\icmltitle{Causal Bandits: Learning Good Interventions via Causal Inference}
% ALTERNATIVES:
% - Causal Bandits: Sequential Decision Making with Causal Models

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2015
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{causal,bandit}

\vskip 0.3in
]

\begin{abstract} 
We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
than algorithms that do not use the additional causal information.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
\input{introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PROBLEM SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Setup}
\label{sec:defs}
\input{setup}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% UPPER BOUND ON SIMPLE REGRET
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Regret Bounds for Parallel Bandit}
\label{sec:simple-regret}

\input{simple-regret}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% UPPER BOUND ON SIMPLE REGRET GENERAL CASE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Regret Bounds for General Graphs}
\label{sec:simple-regret-general}
\input{general}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% EXPERIMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}
\label{sec:experiments}
\input{experiments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion \& Future Work}
\label{sec:discussion}

\input{discussion}

\bibliography{libraryicml}
\bibliographystyle{icml2016}

\ifsup
\onecolumn

\input{tech}
\fi

>>>>

\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

%\setlength{\marginparwidth}{10ex}%added by Yifan so that the todos can fit in the margin
%\usepackage[obeyFinal]{todonotes}
\usepackage[disable]{todonotes}

@@

\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\latent}{\mathbf{L}}
\newcommand{\observed}{\mathbf{O}}
\newcommand{\variables}{\mathbf{V}}

@@

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Identifiability}

We have three questions relating to the general idea of identifiability in causal systems. For all cases, assume we are working with a causal directed acyclic graph. Initially, lets only consider deterministic interventions on single variables in a DAG (not as restrictive as it seems since non-deterministic/multiple interventions can still be modelled by adding nodes to the graph). More formally, we have a causal directed acyclic graph $G$ over variables $\variables = \observed \cup \latent$ where $\observed$ are observable and $\latent$ are latent. Let the set of causal queries $\mathcal{Q}$ be $P(Y|do(X))$ for all possible pairings of variables $X$,$Y$ in $\observed$.

\begin{definition}[Graphical identifiability] a query $Q = P(Y|do(X))$ is identifiable iff we can write an expression for it in terms of conditional probabilities of variables in $\observed$. 
\end{definition}

If a query is graphically identifiable then we can asymptotically obtain the interventional distribution of $Y$ given $do(X=x)$ for any value of $x$ in any probability distribution $P$ compatible with $G$ (with some assumptions on probabilities not being 0). Any desired metric (eg difference in mean, KL divergence) can be applied to the pre and post-interventional distributions to obtain a summary of the \textit{causal effect}.

\subsection*{Partial identifiability}

The graphical identifiability criterion is boolean. There are two directions we can come at trying to smooth this boundary.

\begin{enumerate}
\item Quantify the difficulty in identifying causal effects for problems that are graphically identifiable. What is the variance in the estimators at finite time? What is the minimal change in the graph that makes it non-identifiable?

\item Quantify how well we can identify causal effects for problems that are not graphically identifiable. For example, how tight are the bounds we can get, what is the minimal change in the graph that would make the problem identifiable (eg observe nodes in $\mathcal{U}$ or delete edges),what minimal assumptions must we make about functional relationships between variables to get a good estimate. 
\end{enumerate}

\subsection*{Causal Privacy}

Closely related to the question of partial identifiability. What changes can we make that take a problem from identifiable to non-identifiable.

\subsection*{Characterizing graphical identifiability}

There is an algorithm that determines if a given query in a causal graph is identifiable or not. \ref{} There are also a number of simple graphical criteria that can be used to determine if a graph is or is not identifiable (but no single simple necessary and sufficient criteria). These are:

\begin{itemize}
\item The backdoor criteria (sufficient)
\item The frontdoor criteria (sufficient)
\item (nesseesary)
\end{itemize}

In practise, the back-door criterion is used to estimate causal effects in the vast majority of applications. The front door criterion is also used occasionally.

The question is - can we divide the space of causal queries into sets depending on how they can (or cannot) be identified and estimate the size of the each set? Are there interesting features of graphs that predict identifiability?

A given query can be identified in more than one way. We could try to determine all the ways in which a query could be identified or apply a hierarchy, for example: identifiable by backdoor, identifiable by frontdoor (but not backdoor, identifiable but not via backdoor or frontdoor, not identifiable.

There are three key components to this problem. Identifying a space of graphs, traversing the space in an efficient way and algorithms to compute the quantities we are interested in. There are a very large number of possible causal graphs, even  if we restrict the number of nodes. There are also a number of ways of defining the space of graphs (for example all DAGs versus all MAGs). The existing Identify algorithm \ref{} determines if a query is identifiable or not - and returns an expression if it its. It does not enumerate all the expressions via which a query could be identified. There are algorithms that can determine if a query can be identified via a backdoor criterion \ref{}. No such algorithm that I know of for the front-door criterion. Ideally we want an algorithm that can traverse the space of graphs and efficiently utilize information from the previous (related query) to compute identifiability of the next query.

Related questions - 
\begin{itemize}
\item Are the expressions that the Idenfiy algorithm outputs optimal? Ie - do backdoor criterion adjustments with fewer variables get emmitted in preference to backdoor adjustments with larger numbers of variables? 
\item If there are multiple approaches to identify a given query, how do we define which is optimal? What assumptions do we need to make about the data to define optimality? 
\item How does one efficiently compute causal effects from such expressions \ref{Richardson, etc}?
\end{itemize}

### 2016-06-08 privacy/causal_differential_privacy.tex
adds initial writup of contextual-bandit and identifiability problems

\begin{lemma}
One of the following holds.
\begin{enumerate}
\item $P(Y|do(X)) = P(Y)$
\item There is no set of variables that satisfy the back-door criterion for identifying $P(Y|do(X))$ 
\item There is at least one set of variables $Z$ that are a minimal adjustment for $P(Y|do(X))$, $Z$ may be the empty set. 
\end{enumerate}
\end{lemma}

For the remainder we assume that all queries we wish to jam are of case 3, because in case 1, there is no causal effect of $X$ on $Y$ to discover and in case 2, the causal effect of $X$ on $Y$ can already not be inferred from the data and no way in which we partition the data will render it identifiable.

@@

@@

>>>>

%Causal effects differ from non-causal ones due to bias introduced by confounding variables, which effect both treatment assignment and outcome. For example, ...

Adjusting for nuisance or confounding variables, either by matching or regression, has been a key technique in inferring causal effects at least as far back as \cite{}. The widely used back-door criterion \cite{Pearl2000} clarifies which variables it is appropriate to condition on in order to achieve unbiased estimates of causal effects.

\begin{lemma}Where $X$ and $Y$ are singleton variables, $Z$ is a minimal covariate adjustment for identifying the causal effect of $X$ on $Y$ if and only if $Z$ is a minimal set satisfying the backdoor-criterion relative to $X$ and $Y$ \cite{textor2012adjustment}.
\end{lemma}

The existence of a set covariate adjustment is sufficient but not necessary for identifiability. A causal query can also be identified

We will consider non-parametric point identifiability as described by \cite{Pearl2000}.

The do-calculus provides a complete framework to determine if and how a causal effect may be  identified from observational data. The framework assumes that the data is generated by a directed acyclic graph (DAG) and that the structure of the graph is known (although some variables may not be observed).

is a sufficient criterion for identification of causal effects via matching. More recently it has been generalized to provide necessary and sufficient critera for identification via matching for DAGs \cite{}, and MAGs PAGs \cite{}.

More recently, \cite{Pearl2000} developed the do-calculus, which provides a complete framework for determining if a causal query can be identified from observational data.

\begin{definition}
Identifiability. 
A query $P(Y|do(X))$ is identifiable in a causal graph $G$ if it can be computed uniquely from any positive probability distribution over the observed variables - that is, we can compute an expression for $P(Y|do(X))$ that contains only factors of the observed distribution. 
\end{definition}
\begin{enumerate}
\item $P(Y|do(X)) = P(Y)$. 
\label{cat:noncausal}
\item $P(Y|do(X)) = \sum_Z P(Y|X,Z)P(Z)$
\label{cat:adjust}
\item $P(Y|do(X))$ is identifiable but not via (1) or (2)
\label{cat:identify_other}
\item $P(Y|do(X))$ is not identifiable
\label{cat:not_identifiable}
\end{enumerate}

In this paper we focus on jamming queries that fall into case \ref{cat:adjust}. Case \ref{cat:noncausal} cannot be jammed as the SPCP does not modify marginal distributions. However, it is also unlikely to be sensitive as it holds only if there is no casual relationship between $X$ and $Y$. In case \ref{cat:not_identifiable}, the causal effect cannot be identified from this data set so there is no further work required to jam it. Case \ref{cat:identify_other} is complex and rarely used in empirical work. We leave it for future work.

@@

Note: this holds only if $Z$ is a \textit{minimal} adjustment. If $Z$ is not minimal then we can let $Z = Z' \cup U$ where $Z'$ is a minimal adjustment. In this case, we can still estimate the causal effect of $X$ on $Y$ by conditioning on $Z'$.

@@

\subsection{Further work}

We have considered only single variable interventions and outcomes. We believe the results generalize in a straightforwardly to the case where, for each causal query, $X$ and $Y$ are sets with a minor assumption that there are no causal paths from one intervention variable $X$ to another.

Although overwhelmingly used in practice, adjustment is not necessary for causal identifiability. A given causal graph may also be identifiable via the front door criterion \cite{} or via an expression returned by the identify algorithm \cite{}.

What is the psuedo code to jam all mechanisms? We need an algorithm to produce all 'minimal' expressions. Then each expression will contain a set of factors. At least one of those factors must be jammed for every expression.

Note: A given query may be identifiable via multiple different expressions. As we must jam all of them, it is not sufficient to have an algorithm that returns \textit{an} expression (if one exists). We are not aware of any such algorithms for the general identifiability problem.

Prove that a single query can still always be jammed by separating $X$ and $Y$. Approach  - the expression is a result of the do calculus and nessesarily contains a term with both $X$ and $Y$ together unless...
